\bf Estimating the number of clusters using cross-validation

### 2019-01-11 JCGS-R2/cvclust.tex
twocluster caption

We can consider the problem of choosing $k$, the number of clusters, to be a
model selection problem. In other domains, especially supervised learning
problems like regression and classification, cross-validation is popular for
performing model selection.
In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  We do this
by randomly assigning a fold number $r_i$ in the range $\{1, \dotsc, K\}$
for each row~$i$. For each ``fold'' $r$ in the range $1, \dotsc, K$, we
permute the rows of the data matrix to get $\dataX$, a matrix with the
$r$th test subset, the rows $i$ with $r_i = r$ as its trailing rows.  
We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$, containing roughly $N
(K-1)/K$ of the observations to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{main_code/demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$|\rho| < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{main_code/demo/overlap/color_plot.pdf}
\caption{Number of times $k=2$ is selected out of $10$ replicates for each
pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by
the algorithm, with light (blue) means $k = 1$ is preferred to $k = 2$, and
dark (red) indicates $k = 2$ is preferred to $k = 1$.
The black line is the theoretical boundary determined from
Proposition~\ref{prop:twoclust}.}
\label{fig:overlap-color_plot}
\end{figure}

>>>>

We can consider the problem of choosing $k$, the number of clusters, to be a
model selection problem. In other domains, especially supervised learning
problems like regression and classification, cross-validation is popular for
performing model selection.
In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  We do this
by randomly assigning a fold number $r_i$ in the range $\{1, \dotsc, K\}$
for each row~$i$. For each value $r$ in the range $1, \dotsc, K$, we define
the $r$th ``fold'' as the matrix gotten by permuting the rows of the 
original data matrix to get $\dataX$, a matrix with the
$r$th test subset, the rows $i$ with $r_i = r$ as its trailing rows.  
We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$, containing roughly $N
(K-1)/K$ of the observations to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{main_code/demo/nullcorr/equal.pdf}
\caption{Gabriel cross-validation performance in a single-cluster simulation.
    Gabriel cross-validation error $\CV(k)$ for $k$ ranging from 1 to~5,
    averaged over $10$ replicates where
    data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The 
    criterion chooses the correct answer $k = 1$ whenever
$|\rho| < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{main_code/demo/overlap/color_plot.pdf}
\caption{
    Gabriel cross-validation performance in a two-cluster simulation.
    For a grid of values for $\muX$ and $\muY$ ranging from $0$ to $3$,
    we generated data according the manner described in
    Prop.~\ref{prop:twoclust}, then computed Gabriel cross-validation
    error $\CV(1)$ and $\CV(2)$. The heat map shows the proportion of times
    that $\CV(2) < \CV(1)$ with
    light (blue) indicating that $k = 1$ is preferred to $k = 2$, and
dark (red) indicating that $k = 2$ is preferred to $k = 1$.
    The black line is the theoretical boundary determined from
Proposition~\ref{prop:twoclust}. The simulation confirms our theoretical
    boundary.}
\label{fig:overlap-color_plot}
\end{figure}
